---
output:
  word_document: default
  html_document: default
---
# Alewives

This chapter aims to explore the habitat preferences and life cycle of alewives (Alosa pseudoharengus) in the northeastern United States. Alewives have faced significant declines, leading to their classification as a "species of concern" by the U.S. National Marine Fisheries Service in 2006 (Citation). They have also been considered for inclusion in the U.S. Endangered Species List, as indicated in reports by NOAA Fisheries in 2009 and 2013 (NOAA 2009, 2013). Factors contributing to their decline include diminished water quality, offshore bycatch, and dam construction (Bethoney et al., 2014; Kocovsky et al., 2009).

Recent stock assessment reports reveal diverse trends in documented alewife runs over the last ten years, with some populations showing signs of stabilization or even growth (ASMFC, 2017). Additionally, in 2019, the National Marine Fisheries Service concluded that listing the alewife as threatened or endangered under the Endangered Species Act (ESA) was not warranted (NOAA, 2019).

Alewives are widely distributed throughout the northeastern United States, thriving in freshwater rivers and estuaries along the Atlantic coast (ASMFC, 1985). Historically, they undertook extensive migrations to spawn in freshwater tidal systems, but limited information is available about their estuary and marine movements during the juvenile and adult phases (McCartin et al., 2019).

This chapter aims to explore the favorable habitat conditions for spawning adults, nonmigratory juveniles, and alewife larvae, which are influenced by factors such as suitable spawning habitats, water quality conditions, and availability of appropriate food resources (Lynch et al., 2015).

## Life cycle overview

The alewife exhibits a complex life cycle characterized by distinct stages and behaviors. Spawning typically occurs in waves during the spring season, triggered by rising water temperatures and increasing day length (Able et al., 2020; ASMFC 2009). Adult alewives migrate upstream from their marine environment to reach suitable spawning habitats (Bigelow & Schroeder, 1953). Recent observations show that alewife migration can also be correlated with the lunar phase (Legett et al., 2021).

Upon arrival at the spawning grounds, adult alewives engage in impressive spawning runs, where large aggregations gather to deposit their adhesive eggs over gravel or other hard substrates (Pardue, 1983; Janssen & Luebke, 2004). After spawning, both males and females return to the marine environment, but recent studies suggest that alewives can make multiple spawning trips upstream during the migratory season (Bigelow & Schroeder, 1953; McCartin et al., 2019).

After the spawning process, the eggs go through an incubation period that typically lasts for several days (Bigelow & Schroeder, 1953). Once hatched, the larvae begin their migration upstream, eventually making their way towards estuary habitats where they will reside as they grow (Pardue 1983). This estuary environment serves as a nursery for juvenile alewives until they eventually migrate to the sea (Kosa & Mather, 2001; Laney, 1997). It is noteworthy that the survival rate for larvae is relatively low, with only a small percentage successfully returning to the sea for each female alewife that entered the spawning grounds (Kissil 1974). Similarly, mortality rates for migratory adults during a spawning season can reach as high as 90% in southern regions (Brady et al., 2005). 

## Habitat Requirements 
### Spawning Adult Alewives 

Spawning adult alewives exhibit specific preferences and requirements related to various habitat factors. Their annual migration during spawning is energetically demanding, with notable variations in behavior observed. Some studies report fasting during the day and extensive feeding at night, while others document refraining from eating until their return downstream to productive tidal habitats (Bigelow & Schroeder 1953; Janssen & Brandt 1980). The preferred habitats for spawning are lacustrine and fluvial environments rather than riverine (Reback et al., 2004; Frank et al., 2011; Overton et al., 2012).

Temperature preferences during spawning vary across studies, but there is a consensus that optimal temperatures for successful spawning fall within the range of 12 to 16 degrees Celsius (Brown et al., 2000). Suitable spawning temperatures broadly span from 8 to 22 degrees Celsius (Tyus, 1974; Pardue, 1983; Collette and Klein-MacPhee, 2002; Mather et al., 2012) and ceases above 29 degrees Celsius (Kissil, 1974; Pardue, 1983). Deviations from the optimal temperature range can significantly impact spawning success and the timing of migration. Water temperature also plays a critical role in alewife abundance and movement patterns (Legett et al., 2021).

Regarding depth preferences, spawning adult alewives are generally known to favor depths ranging from MLT-10 meters (Brown et al., 2000), but recent field observations indicate that a significant proportion of alewives can be found in habitats shallower than 2 meters Mather et al. (2012). Alewives are capable of spawning in both shallow and deep water environments, highlighting their adaptability in selecting suitable spawning locations (O'Connell & Angermeier, 1997).

The current understanding of adult alewife spawning behavior challenges the traditional assumption that anadromous species exclusively rely on freshwater environments for reproduction. Brown et al. (2000) found that alewives prefer habitats with salinity concentrations below 15 psu for spawning, while concentrations surpassing 20 psu are deemed unsuitable. However, field studies have documented adult alewives engaging in spawning activities across a diverse array of estuarine habitats with varying salinity levels, including ponds within coastal systems, pond-like regions within coastal rivers and streams, oxbows, eddies, backwaters, stream pools, and flooded swamps (Pardue 1983; Mullen et al. 1986; Collette and Klein-MacPhee 2002; Walsh et al. 2005). This revised understanding challenges the longstanding notion that alewives are exclusively obligated to freshwater for spawning purposes. Additionally, laboratory experiments by DiMaggio et al. (2016) have demonstrated notable survival rates of alewife embryos at salinities ranging up to and including 30‰, indicating their adaptability to a wider range of salinity conditions during their reproductive process.

Flow velocity is a crucial factor influencing the spawning of alewives (Tommasi et al., 2015). These fish are known to prefer slow-moving habitats with little or no current for their spawning activities, as indicated by Walsh et al. (2005). Pardue identifies velocities up to 0.3 m/s as suitable for spawning. However, Haro et al. (2004) conducted experiments showing that migratory alewives can travel farther distances upstream when flow velocities are up to 1.5 m/s, compared to 3.5 m/s. Notably, these experiments reveal very little suitability for spawning around 4.5 m/s (Haro et al., 2004). Understanding the preferred flow velocities is essential in managing and preserving the habitat conditions required for successful alewife spawning.

Previous studies have presented conflicting information regarding the substrate preferences of spawning adult alewives. While some studies suggest that alewives appear to prefer spawning over hard substrates such as gravel and rock (Pardue, Brown et al., 2000), possibly due to the eggs' better adhesion to such surfaces, recent research provides evidence supporting a broader range of substrate utilization by alewives. Notably, Able et al. (2020) documented observations of alewives spawning over sandy substrates, alongside the presence of eggs near hard substrates. Additionally, O’Connell & Angermeier (1997) also support the idea that alewives are known to spawn over a range of substrates, including gravel, sand, vegetation, and other soft substrates. Furthermore, spawning adult alewives do not appear to show a strong preference for habitat containing sub-aquatic vegetation (Rozas & Odum, 1988) (Killgore et al., 1988). Understanding these substrate preferences is crucial for managing and preserving suitable spawning habitats for alewives.

### Premigratory Juvenile Alewives and Larvae

Premigratory juvenile alewives and larvae exhibit distinct habitat preferences and requirements, which play a crucial role in influencing their survival and growth. Several factors influence the abundance and successful development of these young alewives, including river flow, temperature, salinity, depth, and substrate (Pardue, 1983; Walsh et al. 2005; Tommasi et al., 2015). Understanding these ecological factors is vital for effective alewife management, informed decision-making, and the preservation of alewife populations in critical habitats.

Temperature significantly influences the distribution, behavior, and early development of premigratory juvenile alewives and larvae. Optimal temperatures for juvenile alewife development fall within the range of 20°C to 23°C, with a broader suitability range for juvenile and larvae recruitment from 16°C to 26°C (Fay et al., 1983; Tommasi et al., 2015). Hatching success is optimal at around 20.8°C, decreases significantly between 26.7°C to 26.8°C, and ceases entirely above 29.7°C (Pardue, 1983). Juvenile alewives prefer temperatures ranging from 12°C to 30°C, with optimal suitability from 20°C to 22°C and low suitability at 6°C (Pardue, 1983; Brown et al., 2000; Able et al., 2020). Juvenile river herring do not survive temperatures of 3°C or less, rendering such temperatures unsuitable (Otto et al., 1976). Alewife eggs can develop within a temperature range of 11 to 28 degrees Celsius, but hatching success is optimal around 20.8°C, significantly decreases at 26.7°C-26.8°C, and stops completely around 29.7°C (Pardue, 1983; Klauda et al., 1991). Maintaining water temperatures within these ranges is crucial for the successful development and overall health of premigratory juvenile alewives and larvae, as deviations can impact their growth rates and survival.

The depth preferences of premigratory juvenile alewives differ from their adult counterparts, as juveniles exhibit a preference for depths ranging from 0 to 10 meters, with no habitat suitability observed beyond 20 meters (Brown et al., 2000; Höök et al., 2008). Research by Pardue (1983) further supports this finding, indicating that juveniles prefer depths between 0.5 to 5 meters. Studies conducted in Lake Ontario, Canada, revealed that early post-hatch larvae were most abundant in shallow areas less than 3 meters deep, with larger larvae occupying progressively deeper habitats (Ingel, 2013). Additionally, observations in the Margaree River, Nova Scotia, showed that alewife larvae were predominantly found in areas shallower than 2 meters, while the abundance of juvenile alewives increased around five meters deep (Gibson & Myers, 2003). These shallow-water habitats offer protection from predators and facilitate access to suitable food sources, promoting proper growth and development before their downstream migration.

Salinity is another crucial factor influencing the habitat preferences of premigratory juvenile alewives. Research by Turner & Limburg, 2016 and Able et al., 2020 indicates that juvenile alewives typically prefer estuarine habitats with salinity concentrations ranging from 0.5 to 25 psu. These moderate salinity levels provide an optimal balance between freshwater and marine environments, creating suitable conditions for juvenile growth and survival. Higher salinity concentrations, exceeding 20 psu, may limit their habitat suitability, impacting their feeding and physiological processes. Conversely, exposure to higher salinities, up to 30 psu, has been observed to have minimal adverse effects on juvenile alewives' health and survival where 100% survival is shown for 15 psu and decreased survival is experiences around 0 psu and 30 psu (DiMaggio 2015).Understanding the salinity requirements of premigratory juvenile alewives is essential for managing and preserving estuarine habitats that support their early life stages. Maintaining suitable salinity levels in these habitats ensures the healthy development and successful transition of juveniles as they prepare for their migration to the sea. Salinity preferences are higher for juvenile alewives than spawning adults. Juvenile alewives were listed to have a high preference for habitat with salinity concentrations >10 psu but show some suitability for freshwater tidal environments (Brown et al (2000)). Can even be found in concentrations up to 30 psu (Pardue). Fay et al., 1983 reports larvae and juveniles being found in areas with less than 12 psu.

Flow velocity plays a significant role in shaping the development and survival of premigratory juvenile alewives (Tommasi et al., 2015). According to Pardue 1983, When the surface flow ranges from 0 to 0.3 m/s, conditions are deemed suitable for larvae, and egg development. Research by Richkus, 1975 highlights that juvenile alewives show a preference for habitats with flow velocities ranging from 0.05 to 0.17 m/s. O'Connell (1999) proposes that habitat patches with velocity values ranging from 0.06 m/s to 0.16 m/s have a 50% probability of alewife egg presence. Larval alewives were consistently found in water velocities up to approximately 0.12 m/s but not in faster currents (Ingel, 2013). It should be noted that higher flow velocities also have the potential to displace recently spawned eggs from their initial spawning location, as indicated by Able et al 2020. These sluggish flow rates offer suitable conditions for juveniles to conserve energy while efficiently foraging for food. Higher flow velocities may hinder their ability to maintain their position in the water column and access critical food resources (Haro et al., 2004). 

Substrate preferences of premigratory juvenile alewives are diverse, reflecting their adaptability to various environments. Previous studies mention that juvenile alewives show a preference for sandy substrates (Fay et al., 1983), but more recent studies by Able et al., 2020 and Brown et al., 2000 report that juveniles utilize a wide range of substrates (e.g., including sand, gravel, and sub-aquatic vegetation). Such substrate diversity provides juveniles with ample shelter and food sources, supporting their growth and survival during this crucial life stage. Further, observations in Janssen & Luebke, 2004 support that the juvenile alewives may display a preference for rocky substrate. Seagrass coverage plays a vital role in the habitat of premigratory juvenile alewives. Although Ingel, 2013 mentions alewife larvae may avoid areas with aquatic vegetation, research by Laney, 1997 and Smith & Rulifson, 2015 demonstrates that seagrass beds provide essential nursery habitat, offering refuge from predators and abundant food sources. Seagrass beds enhance water quality by stabilizing sediments and promoting nutrient cycling, creating a favorable environment for juvenile alewives to thrive. Vegetated areas are also known to provide important overwintering habitat (Killgore et al., 1988).

## Habitat suitability models

The Alewives Habitat Suitability models, originally developed by Brown et al. (2000) and Pardue (1983), with reliance on similar sources such as Bigelow and Schroeder (1953), possess several limitations that make them inadequate for current applications. Primarily, these models are constructed solely on observations of alewives' daytime behavior, neglecting their significant nocturnal activity patterns. Recent studies have revealed that alewives are primarily active at night, engaging in feeding and exhibiting substantial downstream movement during these nocturnal periods (Janssen & Brandt, 1980; Janssen, 1978; McCartin et al., 2019). Collette & Klein-MacPhee 2002 even notes that mixed gender groups of alewife spawn in stream pools in the evening. Consequently, the exclusive focus on daytime behavior in the existing models fails to capture the true habitat preferences and requirements of alewives, particularly in estuarine and brackish environments.

Furthermore, the current models predominantly consider variables such as temperature, depth, and substrate, while disregarding other crucial factors that significantly influence alewives' habitat selection, including water flow and life stage differences. This limited scope results in incomplete assessments of their habitat suitability. Additionally, inconsistencies and potential inaccuracies arise within the models due to conflicting information regarding the substrate and depth preferences of alewives. These deficiencies undermine the models' effectiveness in predicting the suitability of habitats for alewives.

To address these shortcomings, updated models should encompass a more comprehensive understanding of alewives' behavior, specifically acknowledging their use of estuarine and brackish habitats. These habitats serve as critical areas for alewives, exhibiting relatively high levels of habitat use (McCartin et al., 2019). Incorporating these estuarine and brackish areas into management strategies is of paramount importance to ensure the conservation and successful management of the species. Notably, utilizing estuaries and brackish habitats for spawning may offer energetically favorable conditions for alewives, as it eliminates the need for them to acclimate to complete freshwater environments (DiMaggio et al., 2015). This recognition highlights the significance of incorporating these habitats into conservation efforts and management plans to safeguard the species and support their reproductive success.

### Spawning Adult Alewives

In light of the aforementioned limitations, updated suitability indices for spawning adult alewives have been developed to provide a more comprehensive understanding of their habitat preferences and requirements. These indices, presented in Figure 2, incorporate key factors influencing spawning success, including depth, salinity, temperature, flow velocity, and substrate. Suitability index breakpoints and formulas are listed in Table 1.

The depth suitability index, shown in Figure 2a, indicates the preferred depth ranges for spawning adult alewives. Alewives exhibit higher suitability (index of 1) for depths ranging from 2 to 10 meters. Shallower depths below 2 meters and deeper depths exceeding 10 meters have a reduced suitability index of 0.5.

The salinity suitability index reflects the salinity preferences of spawning adult alewives and is presented in Figure 2b. Alewives show a heightened preference for salinity concentrations below 15 psu, resulting in a suitability index of 1. Salinity concentrations exceeding 20 psu are considered unsuitable, with a suitability index of 0.5.

The temperature suitability index, represented in Figure 2c, depicts the optimal temperature range for successful spawning. Alewives exhibit a high suitability index of 1 within the temperature range of 12 to 16 degrees Celsius. Temperatures above 29.7 degrees Celsius have limited suitability, with a suitability index of 0.5.

The flow velocity suitability index, shown in Figure 2d, indicates the preferred flow velocity range for spawning. Habitat patches with flow velocities ranging from 0.06 to 0.16 m/s have a high suitability index of 1. Velocities below 0.3 m/s are also suitable, albeit with a reduced suitability index of 0.5.

Lastly, the substrate suitability index assesses the preference for different substrate types and is presented in Figure 2e. Alewives show a higher suitability index of 0.8 for hard substrates such as gravel or rock. Sandy substrates have a slightly lower suitability index of 0.6, while other soft substrates exhibit a suitability index of 0.7.

### Juvenile Alewives
- juvenile
Janssen & Luebke, 2004 describes a significant observations of juvenile herring over rocky substrates.
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

### Figures & Tables
```{r, echo= FALSE, include=TRUE, fig.align = "center", fig.cap="Habitat Range of Alewive in the Northeast United States", out.width="80%"}
# Required packages
#library(ggplot2)
#library(sf)

# Read the alewife habitat data (replace "alewife_habitat.shp" with the actual file name and path)
#alewife_habitat <- st_read("alewife_habitat.shp")

# Plot the map
#ggplot() +
#  geom_sf(data = alewife_habitat, fill = "lightblue", color = "black") +
#  coord_sf() +
#  labs(title = "Alewife Habitat Distribution",
#       subtitle = "Northeastern United States",
#       x = "Longitude", y = "Latitude")
```

```{r, echo= FALSE, include=TRUE, fig.align = "center", fig.cap="Spawning Adult Alewife Suitability Breakpoints", out.width="80%"}
library(knitr)

# Create the data for the table
index <- c("a. Depth", "b. Salinity", "c. Temperature", "d. Flow Velocity", "e. Substrate")
breakpoints <- c("Depth ≤ 2 m: 0.5\nDepth > 2 m: 1",
                 "Salinity < 15 psu: 1\nSalinity ≤ 20 psu: 0.5",
                 "Temperature 12-16°C: 1\nTemperature > 16°C & ≤ 29.7°C: 0.5",
                 "Velocity 0.06-0.16 m/s: 1\nVelocity < 0.3 m/s: 0.5",
                 "Hard (gravel or rock): 0.8\nSandy: 0.6\nSoft (other substrates): 0.7\nSuspended sediments: 0.9")

# Create a data frame
df <- data.frame(Index = index, Breakpoints = breakpoints)

# Print the table using kable
kable(df, format = "markdown")
```

```{r, echo= FALSE, include=TRUE, fig.align = "center", fig.cap="Spawning Adult Alewife Suitability Indices", out.width="80%"}
par(mfrow=c(2, 3))

# Depth Suitability Index
depth_values <- seq(0, 30, length.out = 100)
depth_suitability <- ifelse(depth_values <= 2, 0.5, ifelse(depth_values <= 10, 1, 0))
plot(depth_values, depth_suitability, type = "l", xlab = "Depth (m)", ylab = "Suitability Index")
title("Depth Suitability Index")

# Salinity Suitability Index
salinity_values <- seq(0, 30, length.out = 100)
salinity_suitability <- ifelse(salinity_values < 15, 1, ifelse(salinity_values <= 20, 0.5, 0))
plot(salinity_values, salinity_suitability, type = "l", xlab = "Salinity (psu)", ylab = "Suitability Index")
title("Salinity Suitability Index")

# Temperature Suitability Index
temperature_values <- seq(0, 30, length.out = 100)
temperature_suitability <- ifelse(temperature_values >= 12 & temperature_values <= 16, 1, ifelse(temperature_values > 16 & temperature_values <= 29.7, 0.5, 0))
plot(temperature_values, temperature_suitability, type = "l", xlab = "Temperature (°C)", ylab = "Suitability Index")
title("Temperature Suitability Index")

# Flow Velocity Suitability Index
velocity_values <- seq(0, 1, length.out = 100)
velocity_suitability <- ifelse(velocity_values >= 0.06 & velocity_values <= 0.16, 1, ifelse(velocity_values < 0.3, 0.5, 0))
plot(velocity_values, velocity_suitability, type = "l", xlab = "Flow Velocity (m/s)", ylab = "Suitability Index")
title("Flow Velocity Suitability Index")

# Substrate Suitability Index
substrate <- c("Hard (gravel or rock)", "Sandy", "Soft (other substrates)")
substrate_suitability <- c(0.8, 0.6, 0.7)
barplot(substrate_suitability, names.arg = substrate, xlab = "Substrate", ylab = "Suitability Index")
title("Substrate Suitability Index")
```
